 

### 2021-02-01 vertical-thermal-review.Rmd
worked on organization

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].

~~While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory are still unclear.~~ Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy. This will provide a pathway for understanding how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.

(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)

**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance (Fig. `r fig_schematic`), and the implications for understanding forest ecosystem responses and feedbacks to climate change .**

@@

*(KAT: put intra-specific variation first, as we know a lot more about that. Then, add a brief section in inter-specific variation)*

@@

Are traits shaped more by height or light?

@@

*Leaf area / characteristic dimension*:
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)

*Thermal time constant (probably not a lot out there)*
-	Curtis et. al, 2018 – higher in more exposed leaves (heat up slower, cool slower)
-	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?

Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time. It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016]. Typically, at a steady state, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width. Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017]. Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?

Tserej & Feeley 2021 experimentally showed that the rate of leaf warming is negetively correlated with leaf water content per unit leaf area (Lawren mentioned in the meeting as well?)

[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant. 
In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]

*Carotenoids/Xanthophyll cycle pigments*

*Isoprene production*
-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
- Elsa: I think Ty's section on isoprene would be better suited up here

*(Paragraph on intra- vs inter-specific variation)*

- *Are patterns (in traits, metabolism, ecology) driven by tree height or exposure?*

@@

### Distribution of species and traits across the vertical profile

*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).

Ultimately, the way in which vertical gradients influence growth, mortality, and recruitment will shape forest composition across strata. ...

-	*Spring and fall leaf phenology in temperate deciduous forests (Augsburger). 	Mechanism: more moderate conditions (T, wind) in understory (REF).*
*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al. 2000). One potential mechanism is lower temperature/hydraulic stress in understory*
- something on isoprene production?

**Appears to affect community change under warming (Zellweger et al. 2020).**

@@

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
We structure the review around six major questions (Fig. `r fig_schematic`).

@@

@@

#### Morphological traits
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)

#### Biochemical and physiological traits

##### Carotenoids/Xanthophyll cycle pigments

##### Isoprene production
-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
- Elsa: I think Ty's section on isoprene would be better suited up here

#### Leaf lifespan and Deciduousness ("Lifecycle / ecological traits? ")

##### lifespan
##### Deciduousness

*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al. 2000). *
*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).*

*(KAT: a lot of this contnet should probably move up the the intro to this section)*

### Are traits shaped more by height or light?

@@

@@

@@

*(here, it would be good to add some content on contributions to C cycling across the vertical gradient.  If Camille's paper is published ahead of this, that would be the best source. Meakem et al. 2018, New Phyt has a similar analysis for 3 sites.)*
